---
title: "Geospatial Analysis with R"
subtitle: "Mapping ZICTA's ICT Landscape"
author: "Matteo Larrode | International Growth Centre"
date: "2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: false
    code-line-numbers: false
    highlight-style: github
    fig-align: center
    self-contained: true
    footer: "ZICTA R Training Programme — Spatial Data Analysis Session"
    css: |
      .reveal h1, .reveal h2 { color: #1a3a5c; }
      .reveal .slide-number { color: #888; }
      .reveal blockquote { border-left: 4px solid #1a3a5c; padding-left: 1em; color: #444; }
      .zicta-box { background: #eaf2fb; border-left: 4px solid #1a3a5c; padding: 0.8em 1em; border-radius: 4px; margin: 0.5em 0; }
      .warn-box { background: #fef9e7; border-left: 4px solid #f39c12; padding: 0.8em 1em; border-radius: 4px; margin: 0.5em 0; }
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(geodata)
library(leaflet)
library(leaflet.extras)
library(RColorBrewer)
library(viridis)
library(htmltools)

# Load and prepare all data used in slides
zmb_provinces <- st_as_sf(gadm("ZMB", level = 1, path = tempdir()))
coverage  <- read_csv("../../Datasets/zicta_coverage_by_province.csv")
towers    <- read_csv("../../Datasets/zicta_towers.csv")
usaf      <- read_csv("../../Datasets/zicta_usaf_projects.csv")

provinces_mapped <- zmb_provinces |>
  left_join(coverage, by = c("NAME_1" = "province"))

towers_sf <- towers |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

operator_colours <- colorFactor(
  palette = c("#E41A1C", "#377EB8", "#4DAF4A", "#FF7F00"),
  domain  = towers_sf$operator
)

pal_internet <- colorNumeric("plasma", provinces_mapped$internet_penetration_pct,
                             reverse = TRUE)
```

# Geospatial Analysis with R {background-color="#1a3a5c"}

**Mapping ZICTA's ICT Landscape**

ZICTA R Training Programme | Spatial Data Analysis Session

---

## Today's Session

::: columns
::: {.column width="55%"}
**What we will build:**

1. Province-level choropleth maps
2. Interactive operator infrastructure maps
3. Spatial joins to identify coverage gaps
4. A capstone map for your functional group

**By the end of today:**
You will produce publication-ready maps directly usable in ZICTA's Annual Report.
:::

::: {.column width="45%"}
**Session structure:**

| Time | Topic |
|------|-------|
| 0:00 | Spatial data fundamentals |
| 0:45 | Province choropleth |
| 1:45 | Interactive leaflet maps |
| 2:45 | Spatial joins |
| 3:30 | Capstone exercise |
:::
:::

---

# Why Geospatial Analysis? {background-color="#1a3a5c"}

**The gap in ZICTA's reporting**

---

## What ZICTA's Annual Report Currently Shows

> "Internet penetration in rural areas remains significantly lower than urban areas, with notable disparities across provinces."

::: {.zicta-box}
**The data exists.** ZICTA collects coverage rates, subscription counts, and infrastructure locations across all 10 provinces and 116 districts.

**But the spatial dimension is absent** — this information is presented only in tables.
:::

---

## Three Regulatory Questions a Map Can Answer

::: columns
::: {.column width="33%"}
**1. Where are the gaps?**

Which provinces have the lowest internet penetration and mobile coverage?
:::
::: {.column width="33%"}
**2. Is USAF reaching the right places?**

Are Universal Service Access Fund investments targeting the most underserved areas?
:::
::: {.column width="33%"}
**3. Which areas need infrastructure?**

Where are tower density and broadband access both low enough to warrant regulatory intervention?
:::
:::

---

## What We Will Produce Today

:::: columns
::: {.column width="50%"}
**Static maps (ggplot2)**

- Province choropleth of internet penetration
- Faceted comparison of multiple indicators
- Priority province identification
:::
::: {.column width="50%"}
**Interactive maps (leaflet)**

- Zoom, pan, click
- Operator layer toggles
- Hover labels and popup details
:::
::::

> A map is not a decoration. It is an analytical argument about where something is and why it matters.

---

## Prerequisites and Data

**What you already know:** `dplyr`, `ggplot2`, `left_join()`, `pivot_longer()`

**New packages today:**

| Package | Purpose |
|---------|---------|
| `sf` | Read, transform, and analyse spatial data |
| `geodata` | Download Zambia boundaries directly in R |
| `leaflet` | Interactive browser-based maps |

**Datasets (all loaded in setup):**

- `zicta_coverage_by_province.csv` — 10 provinces, 4 indicators
- `zicta_towers.csv` — 72 licensed tower locations, 4 operators
- `zicta_usaf_projects.csv` — 30 USAF project locations and budgets

---

# Getting the Materials: Git and GitHub {background-color="#1a3a5c"}

**Clone the repo — this is how you get the data**

---

## Why Git? The Problem It Solves

::: columns
::: {.column width="50%"}
**Without version control:**

- `analysis_v1.R`
- `analysis_v2.R`
- `analysis_FINAL.R`
- `analysis_FINAL2.R`
- `analysis_ACTUAL_FINAL.R`

→ No one knows which is current.
→ Collaborators overwrite each other's work.
:::
::: {.column width="50%"}
**With git:**

- One file: `analysis.R`
- Full history of every change
- Who changed what, and when
- Multiple people work simultaneously

→ ZICTA analysts share, review, and build on each other's work.
:::
:::

---

## The Three Concepts You Need Today

::: {.zicta-box}
**Repository (repo)** — A folder tracked by git. Contains all your files *plus* the full history of every change ever made. Lives locally on your machine and remotely on GitHub.
:::

::: {.zicta-box}
**Commit** — A named snapshot. "Here is the state of my work, labelled: *Add province choropleth map*." Commits are permanent and reversible.
:::

::: {.zicta-box}
**Push / Pull** — `push` sends your commits to GitHub. `pull` downloads commits others have pushed. The repo on GitHub is the shared reference point.
:::

---

## Getting the Training Materials: `git clone`

Open **Git Bash** (Windows) or **Terminal** (Mac). Run:

```bash
git clone https://github.com/[ZICTA-ORG]/ZICTA_training.git
cd ZICTA_training
```

This downloads everything — scripts, datasets, slides — in one command.

Then open the project in RStudio:

**File → Open Project → `ZICTA_training.Rproj`**

::: {.warn-box}
Always open via the `.Rproj` file — not by opening an `.R` script directly. The project file sets your working directory to the repo root, which makes all `read_csv("Datasets/...")` paths work automatically.
:::

---

## Submitting Your Capstone: The Full Workflow

After the group exercise, commit and push your map to GitHub:

```bash
# 1. Check what has changed
git status

# 2. Stage your group's file(s)
git add Capstone/GroupA_Engineering/GroupA_capstone.R

# 3. Save a named snapshot
git commit -m "GroupA: Engineering tower gap analysis"

# 4. Share it — push to GitHub
git push
```

Your work then appears at `github.com/[ZICTA-ORG]/ZICTA_training` under `Capstone/GroupA_Engineering/`.

::: {.zicta-box}
**One rule:** only touch files inside your group's folder (`Capstone/GroupX_Name/`). This avoids conflicts with other groups working at the same time.
:::

---

# Spatial Data Fundamentals {background-color="#1a3a5c"}

**sf objects, CRS, and your first map**

---

## sf Objects Are Just Data Frames

```{r show-class}
class(zmb_provinces)
```

```{r glimpse-sf, echo=TRUE, eval=FALSE}
glimpse(zmb_provinces)
```

```{r show-geometry, echo=FALSE}
zmb_provinces |>
  select(NAME_1, COUNTRY) |>
  head(3)
```

::: {.zicta-box}
The `geometry` column holds the polygon shapes. Every `dplyr` verb you know — `filter()`, `mutate()`, `left_join()` — works on sf objects. The geometry column is **sticky**: it follows the data even when you don't select it.
:::

---

## What Is a Coordinate Reference System?

::: columns
::: {.column width="50%"}
**EPSG:4326 — WGS84**

- Units: **degrees** of latitude/longitude
- The GPS standard — your phone uses this
- Good for **display and plotting**
- Bad for measuring distances and areas

```r
st_transform(data, crs = 4326)
```
:::
::: {.column width="50%"}
**EPSG:32735 — UTM Zone 35S**

- Units: **metres**
- Optimised for **southern-central Africa**
- Good for **distance and area calculations**
- Required before calling `st_distance()`

```r
st_transform(data, crs = 32735)
```
:::
:::

---

## Why CRS Matters: The Area Trap

```{r crs-demo}
zmb_provinces |>
  mutate(
    area_wrong = as.numeric(st_area(geometry)) / 1e6,
    area_sqkm  = as.numeric(st_area(st_transform(geometry, 32735))) / 1e6
  ) |>
  st_drop_geometry() |>
  select(NAME_1, area_wrong, area_sqkm) |>
  arrange(desc(area_sqkm)) |>
  head(5)
```

::: {.warn-box}
`area_wrong` is in degrees-squared — meaningless. Always `st_transform()` to EPSG:32735 before measuring in Zambia.
:::

---

## Downloading Zambia Boundaries in R

```{r gadm-demo, eval=FALSE}
library(geodata)

# Download directly — no shapefile download needed
zmb_provinces_raw <- gadm("ZMB", level = 1, path = tempdir())
zmb_districts_raw <- gadm("ZMB", level = 2, path = tempdir())

# Convert SpatVector → sf (required for ggplot2 and leaflet)
zmb_provinces <- st_as_sf(zmb_provinces_raw)
zmb_districts <- st_as_sf(zmb_districts_raw)
```

::: {.zicta-box}
`gadm()` returns a **SpatVector** (terra package). Always call `st_as_sf()` immediately after. Forgetting this is the #1 error when using `geodata`.
:::

Province names in `NAME_1`: `r paste(sort(zmb_provinces$NAME_1), collapse = ", ")`

---

## Your First Regulatory Map

```{r first-map, fig.height=4.5, fig.width=7}
ggplot(zmb_provinces) +
  geom_sf(fill = "#f0f0f0", colour = "grey40", linewidth = 0.4) +
  geom_sf_label(aes(label = NAME_1), size = 2.2, fill = "white",
                label.size = 0, label.padding = unit(0.1, "lines")) +
  labs(title = "Zambia — 10 Provinces",
       caption = "Source: GADM (geodata package)") +
  theme_void()
```

---

# Province Choropleth {background-color="#1a3a5c"}

**Connecting ZICTA data to the map**

---

## The Core Pattern: A Spatial Attribute Join

This is `left_join()` — exactly what you learned in Week 10.
The only difference: the left table is an sf object, so the result is also spatial.

```{r join-demo}
provinces_mapped <- zmb_provinces |>
  left_join(coverage, by = c("NAME_1" = "province"))

# Always validate: NAs here = name mismatch
provinces_mapped |>
  st_drop_geometry() |>
  filter(is.na(internet_penetration_pct)) |>
  pull(NAME_1)
```

::: {.warn-box}
`character(0)` = no mismatches. Any province names appearing here means a spelling mismatch to fix. "North-Western" (with hyphen) is the GADM spelling — your CSV must match exactly.
:::

---

## Basic Choropleth

```{r basic-choro, fig.height=4, fig.width=7}
ggplot(provinces_mapped) +
  geom_sf(aes(fill = internet_penetration_pct)) +
  scale_fill_viridis_c(name = "Internet\npenetration (%)") +
  labs(title = "Internet Penetration by Province, Zambia 2022",
       caption = "Source: ZICTA 2022 | Boundaries: GADM") +
  theme_void()
```

---

## Publication-Ready: Labels and Palette Direction

```{r pub-choro, fig.height=4.2, fig.width=8, echo=FALSE}
province_labels <- zmb_provinces |>
  st_transform(32735) |>
  mutate(centroid = st_centroid(geometry)) |>
  st_drop_geometry() |>
  rename(geometry = centroid) |>
  st_as_sf(crs = 32735) |>
  st_transform(4326)

ggplot() +
  geom_sf(data = provinces_mapped, aes(fill = internet_penetration_pct),
          colour = "white", linewidth = 0.4) +
  geom_sf_text(data = province_labels, aes(label = NAME_1),
               size = 2.1, colour = "white", fontface = "bold") +
  scale_fill_viridis_c(name = "Internet\npenetration (%)", option = "plasma",
                       direction = -1, labels = function(x) paste0(x, "%")) +
  labs(title    = "Internet Penetration by Province, Zambia 2022",
       subtitle = "Lusaka (67.8%) and Copperbelt (52.1%) far exceed rural provinces",
       caption  = "Source: ZICTA 2022 ICT Indicators | Boundaries: GADM") +
  theme_void(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(colour = "grey40"),
        legend.position = "right")
```

`direction = -1` reverses the palette: **darker = lower penetration**, drawing attention to underserved areas.

---

## Colour Palette Choices

::: columns
::: {.column width="50%"}
**`scale_fill_viridis_c()`**

- Perceptually uniform
- Accessible to colour-blind readers
- Options: `"viridis"`, `"plasma"`, `"magma"`, `"cividis"`
- `direction = -1` reverses the scale

Best for: continuous data like penetration rates
:::
::: {.column width="50%"}
**`scale_fill_distiller()`**

- Diverging palettes: `"RdYlGn"`, `"PuOr"`
- Sequential palettes: `"Blues"`, `"Reds"`

Best for: compliance scores, tower counts, data with a natural midpoint

**Rule:** Always explain your palette choice in the map caption.
:::
:::

---

## Faceted Choropleth: Comparing Two Indicators

```{r facet-choro, fig.height=4, fig.width=9, echo=FALSE}
provinces_long <- provinces_mapped |>
  select(NAME_1, internet_penetration_pct, mobile_coverage_pct, geometry) |>
  pivot_longer(cols = c(internet_penetration_pct, mobile_coverage_pct),
               names_to = "indicator", values_to = "value") |>
  mutate(indicator = recode(indicator,
    "internet_penetration_pct" = "Internet Penetration (%)",
    "mobile_coverage_pct"      = "Mobile Network Coverage (%)"))

ggplot(provinces_long) +
  geom_sf(aes(fill = value), colour = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "plasma", direction = -1, name = "Value (%)") +
  facet_wrap(~ indicator) +
  labs(title = "Digital Connectivity Indicators by Province, Zambia 2022",
       caption = "Source: ZICTA 2022 | Boundaries: GADM") +
  theme_void(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), legend.position = "bottom")
```

---

## Reading the Faceted Map

::: {.zicta-box}
**Key observation:** Mobile coverage is universally higher than internet penetration — often by 30–50 percentage points in rural provinces.

Zambia has **towers** (signal exists) but limited **internet use** (service is not being adopted at the same rate).
:::

**What explains this gap?**

- Handset affordability
- Data costs relative to income
- Digital literacy
- Relevance of available content

> This is a **ZICTA policy question**, not just a data observation. The map makes it visible.

---

# Interactive Maps with leaflet {background-color="#1a3a5c"}

**Zoom, click, and toggle — maps for exploration**

---

## leaflet vs ggplot2

| | ggplot2 | leaflet |
|--|---------|---------|
| Output | Static PNG/PDF | Interactive HTML |
| Best for | Reports, slides, publications | Dashboards, exploration, stakeholder meetings |
| Syntax | `+` to add layers | `%>%` to add layers |
| Palette functions | `scale_fill_viridis_c()` | `colorNumeric()` |
| Data CRS required | Any (ggplot handles it) | **Must be EPSG:4326** |

::: {.warn-box}
If your leaflet map appears blank, the most common cause is data in a projected CRS (e.g., 32735). Always `st_transform(4326)` before passing to leaflet.
:::

---

## The leaflet Pipeline

```{r leaflet-basic, eval=FALSE}
pal_internet <- colorNumeric(      # ① Build colour palette
  palette = "plasma",
  domain  = provinces_mapped$internet_penetration_pct,
  reverse = TRUE
)

leaflet(provinces_mapped) %>%      # ② Initialise with data
  addProviderTiles("CartoDB.Positron") %>%   # ③ Base map tiles
  addPolygons(                               # ④ Add polygon layer
    fillColor   = ~pal_internet(internet_penetration_pct),
    fillOpacity = 0.75,
    color       = "white",
    weight      = 1
  ) %>%
  addLegend(                                 # ⑤ Add legend
    pal    = pal_internet,
    values = ~internet_penetration_pct,
    title  = "Internet Penetration (%)"
  )
```

The `~` (tilde) means **"evaluate as a formula on the data"**. You will see it in every `fillColor`, `label`, and `popup` argument.

---

## Interactive Choropleth with Hover Labels

```{r leaflet-hover, echo=FALSE}
leaflet(provinces_mapped) |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    fillColor   = ~pal_internet(internet_penetration_pct),
    fillOpacity = 0.75,
    color       = "white", weight = 1.5,
    highlight   = highlightOptions(weight = 3, color = "#FFD700", bringToFront = TRUE),
    label = ~paste0(
      "<b>", NAME_1, "</b><br>",
      "Internet: <b>", internet_penetration_pct, "%</b><br>",
      "Mobile coverage: ", mobile_coverage_pct, "%"
    ) |> lapply(htmltools::HTML)
  ) |>
  addLegend(pal = pal_internet, values = ~internet_penetration_pct,
            title = "Internet (%)", position = "bottomright")
```

---

## Converting a CSV to Spatial Points

```{r st-as-sf-demo}
towers_sf <- towers |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
```

::: {.warn-box}
**Order matters:** `coords = c("longitude", "latitude")` — X (longitude) **before** Y (latitude). Swapping these places all points in the southern Atlantic. R does **not** warn you.
:::

```{r check-points, fig.height=2.5, fig.width=5, echo=FALSE}
par(mar = c(2, 2, 1, 1))
plot(st_geometry(zmb_provinces), col = "#f5f5f5", border = "grey60")
plot(st_geometry(towers_sf), add = TRUE, col = "#E41A1C", pch = 16, cex = 0.8)
title("Tower locations — quick sanity check", cex.main = 0.9)
```

---

## Tower Map with Operator Colours

```{r tower-map, echo=FALSE}
leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(data = provinces_mapped, fillColor = "transparent",
              color = "grey60", weight = 1, label = ~NAME_1) |>
  addCircleMarkers(
    data = towers_sf, radius = 6,
    color = ~operator_colours(operator),
    fillColor = ~operator_colours(operator),
    fillOpacity = 0.85, stroke = FALSE,
    popup = ~paste0("<b>", operator, "</b><br>",
                   district, ", ", province, "<br>Year: ", year_installed)
  ) |>
  addLegend(pal = operator_colours, values = towers_sf$operator, title = "Operator")
```

---

## Layered Map: Choropleth + Operators + Toggles

```{r layered-map, echo=FALSE}
leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(data = provinces_mapped,
    fillColor = ~pal_internet(internet_penetration_pct),
    fillOpacity = 0.55, color = "white", weight = 1.2,
    label = ~paste0(NAME_1, ": ", internet_penetration_pct, "%") |> lapply(htmltools::HTML),
    group = "Internet Penetration"
  ) |>
  addCircleMarkers(data = towers_sf |> dplyr::filter(operator == "Airtel Zambia"),
    radius = 5, color = "#E41A1C", fillColor = "#E41A1C",
    fillOpacity = 0.9, stroke = FALSE,
    popup = ~paste0("<b>", operator, "</b><br>", district, ", ", province),
    group = "Airtel Zambia"
  ) |>
  addCircleMarkers(data = towers_sf |> dplyr::filter(operator == "MTN Zambia"),
    radius = 5, color = "#377EB8", fillColor = "#377EB8",
    fillOpacity = 0.9, stroke = FALSE,
    popup = ~paste0("<b>", operator, "</b><br>", district, ", ", province),
    group = "MTN Zambia"
  ) |>
  addCircleMarkers(data = towers_sf |> dplyr::filter(operator == "ZAMTEL"),
    radius = 5, color = "#4DAF4A", fillColor = "#4DAF4A",
    fillOpacity = 0.9, stroke = FALSE,
    popup = ~paste0("<b>", operator, "</b><br>", district, ", ", province),
    group = "ZAMTEL"
  ) |>
  addCircleMarkers(data = towers_sf |> dplyr::filter(operator == "Liquid Telecom"),
    radius = 5, color = "#FF7F00", fillColor = "#FF7F00",
    fillOpacity = 0.9, stroke = FALSE,
    popup = ~paste0("<b>", operator, "</b><br>", district, ", ", province),
    group = "Liquid Telecom"
  ) |>
  addLayersControl(
    overlayGroups = c("Internet Penetration", "Airtel Zambia", "MTN Zambia",
                      "ZAMTEL", "Liquid Telecom"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

---

# Spatial Joins {background-color="#1a3a5c"}

**Using geography as the join key**

---

## Attribute Join vs Spatial Join

::: columns
::: {.column width="50%"}
**Week 10: Attribute join**

```r
# Shared key column required
left_join(coverage,
  by = c("NAME_1" = "province"))
```

Join key = a column value (province name)
:::
::: {.column width="50%"}
**Today: Spatial join**

```r
# No shared column needed
towers_sf |>
  st_join(zmb_provinces)
```

Join key = **geographic location**
Point falls inside polygon → inherits its attributes
:::
:::

::: {.zicta-box}
`st_join()` is the spatial equivalent of a left join. Each point gets the attributes of whichever polygon it falls inside. If a point is outside all polygons, it gets `NA`.
:::

---

## Validating a Spatial Join

```{r st-join-demo}
towers_enriched <- towers_sf |>
  st_join(zmb_provinces |> select(NAME_1))

towers_enriched |>
  st_drop_geometry() |>
  select(tower_id, operator, province, NAME_1) |>
  mutate(match = province == NAME_1) |>
  count(match)
```

::: {.zicta-box}
All 72 towers correctly fall within their named provinces. In real ZICTA data this check catches GPS transcription errors in infrastructure registries — a common data quality issue.
:::

---

## Aggregating Points to Province Level

```{r agg-points}
towers_per_province <- towers_enriched |>
  st_drop_geometry() |>
  count(NAME_1, name = "n_towers") |>
  rename(province = NAME_1)

provinces_towers <- zmb_provinces |>
  left_join(towers_per_province, by = c("NAME_1" = "province")) |>
  mutate(n_towers = replace_na(n_towers, 0))
```

::: {.warn-box}
`replace_na(n_towers, 0)` is essential. Provinces with **no towers** still need a value to appear correctly on the map. An `NA` would make them transparent or cause an error.
:::

---

## Coverage Gap Analysis: The Key Output

```{r gap-map, fig.height=4, fig.width=7, echo=FALSE}
coverage_gap <- provinces_towers |>
  left_join(coverage, by = c("NAME_1" = "province")) |>
  mutate(priority_flag = internet_penetration_pct < 20 & n_towers < 8)

ggplot(coverage_gap) +
  geom_sf(aes(fill = priority_flag), colour = "white", linewidth = 0.4) +
  scale_fill_manual(values = c("FALSE" = "#d4e6f1", "TRUE" = "#c0392b"),
                    labels  = c("FALSE" = "Not priority", "TRUE" = "Priority (low penetration + few towers)"),
                    name    = "") +
  geom_sf_text(aes(label = NAME_1), size = 2.1, colour = "grey20") +
  labs(title   = "USAF Priority Provinces: Low Penetration and Low Tower Density",
       caption = "Priority: internet penetration < 20% AND fewer than 8 licensed towers") +
  theme_void() +
  theme(legend.position = "bottom")
```

---

## Distance Calculations with st_distance()

```{r distance-demo, eval=FALSE}
# Step 1: Compute the reference point in a projected CRS
lusaka_centroid <- zmb_provinces |>
  filter(NAME_1 == "Lusaka") |>
  st_transform(32735) |>          # ← MUST be projected for distance
  st_centroid() |>
  st_geometry()

# Step 2: Transform towers to the same CRS, then measure
towers_utm <- towers_sf |>
  st_transform(32735) |>
  mutate(dist_lusaka_km = as.numeric(
    st_distance(geometry, lusaka_centroid)
  ) / 1000)
```

::: {.warn-box}
`st_distance()` with geographic coordinates (4326) returns values in degrees — meaningless and silently wrong. **Always project to 32735 first.**
:::

---

# Capstone Exercise {background-color="#1a3a5c"}

**Your group, your map, your regulatory mandate**

---

## Capstone: How It Works

**90 minutes | Group work | 2-minute presentation per group**

Each functional group produces **one publication-ready map** that could appear in the ZICTA Annual Report.

Use the starter code in `ZICTA_Geo_Student.R` Section 5 for your group.
The facilitator will circulate every 10–15 minutes.

::: {.zicta-box}
**What makes a map publication-ready?**

- Clear title and subtitle stating the finding, not just the topic
- Source and date in the caption
- Meaningful colour palette (explained or self-evident)
- Province or district labels where space allows
:::

---

## Group A — Engineering

**Task: ZAMTEL infrastructure gap analysis**

1. Filter the tower registry to ZAMTEL only
2. Count ZAMTEL towers per province and produce a choropleth
3. Flag towers more than 500 km from Lusaka as "remote" on a leaflet map
4. **Findings to present:** Which provinces have fewer than 2 ZAMTEL towers? How many towers are classified as remote?

*Key functions:* `filter()`, `st_join()`, `count()`, `st_distance()`, `addCircleMarkers()`

---

## Groups B & C — Statistical Unit & Financial Statistics

**Group B: Four-indicator comparison + inequality quantification**

1. Add a `towers_per_million` column to the province data
2. Produce a 2×2 faceted choropleth of all 4 connectivity indicators
3. Calculate the max/min ratio for each indicator
4. **Finding to present:** Which indicator shows the greatest provincial inequality?

**Group C: Compliance + penetration relationship**

1. Load `sample_data_telecoms.csv` and compute average compliance by province
2. Map compliance scores as a choropleth
3. Build a scatter plot: compliance score vs internet penetration
4. **Finding to present:** Is there a relationship? What does it suggest for enforcement priority?

---

## Groups D & E — Postal Services & Universal Access

**Group D: USAF project map**

1. Convert `zicta_usaf_projects.csv` to sf with `st_as_sf()`
2. Build a leaflet map coloured by project status (Completed/Ongoing/Planned)
3. Make circle size proportional to budget
4. **Finding to present:** Which province has received the most total USAF investment?

**Group E: Underinvestment targeting**

1. Join USAF budget totals to the province spatial layer
2. Flag provinces where: internet penetration < 20% **AND** total USAF investment < $500k
3. Produce a final priority map and summary table
4. **Finding to present:** Which provinces are priority targets and what is the total population at risk?

---

# Closing {background-color="#1a3a5c"}

---

## Key Functions Reference

| Function | Package | Purpose |
|----------|---------|---------|
| `st_as_sf(coords=...)` | sf | CSV with lat/lon → spatial object |
| `st_transform(crs=...)` | sf | Change coordinate reference system |
| `st_crs()` | sf | Check current CRS |
| `st_join()` | sf | Spatial join (point-in-polygon) |
| `st_centroid()` | sf | Compute polygon centroid |
| `st_distance()` | sf | Distance between geometries |
| `geom_sf()` | ggplot2 | Draw sf objects |
| `geom_sf_text()` | ggplot2 | Labels at geometry locations |
| `colorNumeric()` | leaflet | Continuous colour palette |
| `colorFactor()` | leaflet | Categorical colour palette |
| `addPolygons()` | leaflet | Polygon layer |
| `addCircleMarkers()` | leaflet | Point layer |
| `addLayersControl()` | leaflet | Toggle controls |

---

## Zambia Spatial Data Sources

| Source | What it provides | Access |
|--------|-----------------|--------|
| `geodata::gadm("ZMB")` | Province + district boundaries | Free, in-R |
| OCHA HDX | Admin boundaries, GeoJSON | data.humdata.org |
| WorldPop | Population grid (100m) | worldpop.org |
| Zambia Statistics Agency | 2022 Census tables | zamstats.gov.zm |
| GSMA Coverage API | 2G/3G/4G coverage polygons | gsma.com/coverage |
| OpenCelID | Crowdsourced tower locations | opencellid.org |
| Natural Earth | Country borders, roads, rivers | naturalearthdata.com |

Full annotated list: `Zambia_Spatial_Data_Sources.md` in the session folder.

---

## What Comes Next

::: columns
::: {.column width="50%"}
**Immediate applications**

- Embed today's leaflet maps directly into a flexdashboard (Module 15–16 skills)
- Add these maps to the ZICTA Annual Report template
- Update with real ZICTA data — the code structure is identical
:::
::: {.column width="50%"}
**Extensions to explore**

- District-level analysis: change `level = 1` → `level = 2` in `gadm()`
- Raster data: population grids with `terra`
- OpenStreetMap data: roads, schools via `osmdata`
- Buffer zones: `st_buffer()` for signal coverage approximation
:::
:::

::: {.zicta-box}
All R scripts, datasets, and this slide deck are in `ZICTA_training/Modules/Geo Session/`. The code runs unchanged once the ZICTA datasets are updated with real values.
:::

---

## Thank You

**Matteo Larrode | International Growth Centre**

Session materials:
- `ZICTA_Geo_Live_Script.R` — full annotated facilitator script
- `ZICTA_Geo_Student.R` — fill-in-blank student version
- `Zambia_Spatial_Data_Sources.md` — curated data sources
- `Geo_Session_Facilitation_Guide.md` — re-delivery guide for ZEL

> Questions? Challenges? Ideas for your Annual Report?
